---
title: "Learning Single Cell RNASeq with Seurat: Part 1"
description: "Learning about the single cell RNASeq workflow using the Seurat package and Crohn's Disease dat from the Broad Institute Single Cell Portal"
author: "Sadie Wisotsky"
date: "6/12/2026"
categories:
  - RNASeq
  - scRNASeq
  - analysis
  - R
  - Seurat
format: html
---

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(here)
```

# Part 1: Basic workflow on Colon Stromal data

I wanted to learn about single cell RNASeq and the Seurat package so I have chosen a data set that contains scRNASeq data from multiple tissue sites and compartments. This tissue has already been analyzed in this paper but I am hoping to use that as a check and as way to learn why they chose specific parameters and workflows. 

For this first part, I will be following the basic workflow laid out in the Seurat tutorial for the Colon Stromal data. Further parts will expand out to the other 5 data sets and comparisons between them. 

## Step 1: Retrieve the data

I downloaded the data from the Broad Institute Single Cell portal [here](https://singlecell.broadinstitute.org/single_cell/study/SCP1884/human-cd-atlas-study-between-colon-and-terminal-ileum#study-download). 

There are 6 main data subsets for the expression data. Samples from the colon and the ileum and then broken down by compartment by stromal, epithelial, and immune cells. For this first post, I will be focusing on the colon stromal data.

I wrote a function to create a custom Seurat object which will make it easier and more repeatable to read in the data for all 6 subsets.


```{r}
create_seurat_obj <- function(path) {
   matrix_path <- list.files(path, ".*raw.mtx", full.names = TRUE)
   barcodes_path <- list.files(path, ".*.barcodes.tsv", full.names = TRUE)
   features_path <-  list.files(path, ".*.features.tsv", full.names = TRUE)
  
   sample_name <- matrix_path |> basename()  |> substr(1,6)

   counts_matrix <- ReadMtx(
     mtx = matrix_path,
     cells = barcodes_path,
     features = features_path,
     cell.column = 1,     # Usually barcodes are in the first column
     feature.column = 2   # Change to 1 if your gene file only has 1 column (IDs instead of Symbols)
   )
  # Create the Seurat object
   obj <- CreateSeuratObject(counts = counts_matrix, project = sample_name)
  
  # Tag with custom metadata
   obj$Tissue <- sample_name
  
   return(obj)
}
```

I then create the Seurat object and look at what it is:
```{r}
CO_STR <- create_seurat_obj(here("./data/SCP1884/expression/62a0c9f523a28233e1c9b06b/"))
CO_STR
glimpse(CO_STR)
```

Can see we have 39433 samples and 28663 features. Seurat as of v5.0 allows you to store different assays in layers. Here we can see that our only layer for now is counts and our assay is RNA. I probably won't dive deeper into this in this part but there is a [vignette here](https://satijalab.org/seurat/articles/seurat5_essential_commands) for those interested. 

```{r}
Layers(CO_STR)
```

## Step 2: QC

the first part of pre-processing the data is quality control. I will be following the guidelines outlined in the tutorial for now. Starting with calculating the percentage of mitochondrial (mt) contamination; a high value here is indictive of dying or low quality cells that we want to exclude from our analysis. 

```{r}
CO_STR[["percent.mt"]] <- PercentageFeatureSet(CO_STR, pattern = "^MT-")
```

I peak at the first 5.
```{r}
# Show QC metrics for the first 5 cells
head(CO_STR@meta.data, 5)
```

Then it is time to graph the metrics and visually check if our standard QC filters are appropriate:

```{r}
# Visualize QC metrics as a violin plot
VlnPlot(CO_STR, features = c("nCount_RNA"), group.by="orig.ident")
```

here's where I realize that I am doing this across multiple samples where the tutorial is across a single sample. For me, it is hard to see where the cutoff for number of features should be based on this so let's try to visualize it another way.

```{r}
CO_STR@meta.data %>% 
    ggplot(aes(color=orig.ident, x=nCount_RNA, fill= orig.ident)) + 
    geom_density(alpha = 0.2) + 
    theme_classic() +
    scale_x_log10() + 
    geom_vline(xintercept = 450,color="red",linetype="dotted")+ 
    geom_vline(xintercept = 3500,color="red",linetype="dotted")+ 
    geom_vline(xintercept = 15000,color="red",linetype="dotted")

```



## Step 3: Patient-Specific QC Filtering

Looking at the density plots, it's clear that cells from different patients have different distributions of RNA count and feature count. Using a single global threshold might discard high-quality cells from some samples while keeping low-quality cells from others. 

To address this, I will split the Colon Stromal object by patient (`orig.ident`), and apply customized thresholds for patients that show outlier characteristics (as identified in my exploratory work), while using a robust default threshold for the others.

The outliers and their thresholds:
- **I124246**: High overall features/counts. Threshold: `min_feat = 500`, `max_feat = 6000`, `max_count = 25000`, `max_mt = 8`
- **N104689**: High overall features/counts. Threshold: `min_feat = 500`, `max_feat = 5500`, `max_count = 20000`, `max_mt = 10`
- **H106265**: Low overall features/counts. Threshold: `min_feat = 200`, `max_feat = 2000`, `max_count = 8000`, `max_mt = 5`
- **Default (all other patients)**: `min_feat = 200`, `max_feat = 3000`, `max_count = 12000`, `max_mt = 5`

```{r}
library(purrr)

# Split by patient identity
stromal_list <- SplitObject(CO_STR, split.by = "orig.ident")

# Define tailored patient-specific thresholds
patient_thresholds <- list(
  "I124246" = list(min_feat = 500, max_feat = 6000, max_count = 25000, max_mt = 8),
  "N104689" = list(min_feat = 500, max_feat = 5500, max_count = 20000, max_mt = 10),
  "H106265" = list(min_feat = 200, max_feat = 2000, max_count = 8000,  max_mt = 5),
  # Additional tricky patients identified from summary statistics
  "I121881" = list(min_feat = 200, max_feat = 6000, max_count = 28000, max_mt = 12),
  "I117351" = list(min_feat = 200, max_feat = 6000, max_count = 31000, max_mt = 15),
  "N107306" = list(min_feat = 200, max_feat = 6000, max_count = 30000, max_mt = 15),
  "N154787" = list(min_feat = 200, max_feat = 6000, max_count = 20000, max_mt = 12),
  "N128208" = list(min_feat = 200, max_feat = 6000, max_count = 28000, max_mt = 8),
  "N128624" = list(min_feat = 200, max_feat = 6000, max_count = 25000, max_mt = 8),
  "N105446" = list(min_feat = 500, max_feat = 6000, max_count = 25000, max_mt = 15)
)
default_thresh <- list(min_feat = 200, max_feat = 3000, max_count = 12000, max_mt = 5)

# Filter each patient individually
filtered_stromal_list <- imap(stromal_list, function(obj, patient_id) {
  # Fetch the custom threshold for this patient, OR use the default if not listed
  thresh <- patient_thresholds[[patient_id]]
  if (is.null(thresh)) {
    thresh <- default_thresh
  }
  
  # Calculate MT%
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  
  # Identify cells passing thresholds
  keep_cells <- colnames(obj)[
    obj$nFeature_RNA > thresh$min_feat & 
    obj$nFeature_RNA < thresh$max_feat & 
    obj$nCount_RNA < thresh$max_count &
    obj$percent.mt < thresh$max_mt
  ]
  
  # If no cells pass, return NULL to avoid subsetting error
  if (length(keep_cells) == 0) {
    warning(paste("Patient", patient_id, "has no cells passing thresholds and will be excluded."))
    return(NULL)
  }
  
  # Apply tailored filtering
  obj <- subset(obj, cells = keep_cells)
  return(obj)
})

# Remove NULL elements
filtered_stromal_list <- compact(filtered_stromal_list)

# Merge back into a single clean Seurat object
CO_STR_clean <- merge(
  x = filtered_stromal_list[[1]], 
  y = filtered_stromal_list[-1],
  project = "CO_STR_Clean"
)

# Print comparison of cell counts
cat("Original cells:", ncol(CO_STR), "\n")
cat("Filtered cells:", ncol(CO_STR_clean), "\n")
```

## Step 4: Normalization & Variable Feature Selection

Next, I normalize the counts to control for sequencing depth and find the 2,000 most highly variable genes across cells, which will be used for linear dimensionality reduction (PCA).

```{r}
# Standard log-normalization
CO_STR_clean <- NormalizeData(CO_STR_clean, normalization.method = "LogNormalize", scale.factor = 10000)

# Find variable features
CO_STR_clean <- FindVariableFeatures(CO_STR_clean, selection.method = "vst", nfeatures = 2000)

# Identify top 10 variable features and plot
top10 <- head(VariableFeatures(CO_STR_clean), 10)
plot1 <- VariableFeaturePlot(CO_STR_clean)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Step 5: Scaling & PCA

I scale the data so that each gene has a mean expression of 0 and a variance of 1 across all cells. This ensures that highly expressed genes do not dominate the PCA. Then, I run PCA on the variable features.

```{r}
CO_STR_clean <- ScaleData(CO_STR_clean)
CO_STR_clean <- RunPCA(CO_STR_clean, npcs = 30, verbose = FALSE)

# Visualize PCA load and elbow plot
ElbowPlot(CO_STR_clean, ndims = 30)
```

## Step 6: Batch Integration using Harmony

Since our cells come from 38 different patient samples, we are very likely to observe patient-specific clusters due to technical variations or individual genetic baselines (batch effects). To align equivalent cell populations across patients, I will integrate the datasets using the **Harmony** integration method.

In Seurat v5, this is streamlined using the `IntegrateLayers` function.

```{r}
library(harmony)

# Run Harmony Integration
CO_STR_clean <- IntegrateLayers(
  object = CO_STR_clean, 
  method = HarmonyIntegration,
  orig.reduction = "pca", 
  new.reduction = "harmony",
  verbose = FALSE
)
```

## Step 7: Clustering & UMAP Visualization

With the integrated representation (`harmony` reduction) ready, I will build the Shared Nearest Neighbor (SNN) graph, find clusters, and run UMAP for visualization. I'll use 20 dimensions of the integrated space.

```{r}
# Build SNN Graph and Find Clusters on integrated space
CO_STR_clean <- FindNeighbors(CO_STR_clean, reduction = "harmony", dims = 1:20)
CO_STR_clean <- FindClusters(CO_STR_clean, resolution = 0.5, verbose = FALSE)

# Run UMAP
CO_STR_clean <- RunUMAP(CO_STR_clean, reduction = "harmony", dims = 1:20)
```

Let's visualize the results! I will plot the cells colored by cluster and patient identity side-by-side using the `patchwork` library. This allows us to verify if Harmony successfully mixed the patient samples while preserving biological differences.

```{r}
library(patchwork)

p1 <- DimPlot(CO_STR_clean, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + 
  ggtitle("Integrated UMAP: Clusters") + 
  theme_minimal()

p2 <- DimPlot(CO_STR_clean, reduction = "umap", group.by = "orig.ident", label = FALSE) + 
  ggtitle("Integrated UMAP: Patients") + 
  theme_minimal() +
  theme(legend.position = "none") # Hide large patient legend

p1 + p2
```

We can see that cells from different patients are well-mixed within the same clusters, indicating that the Harmony integration successfully corrected for batch effects! In the next parts, we will extend this pipeline to the other tissue compartments (epithelial and immune) and dive deeper into biological characterization and marker identification.